---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

# autodiffr for Automatic Differentiation in R through Julia

[![Travis-CI Build Status](https://travis-ci.org/Non-Contradiction/autodiffr.svg?branch=master)](https://travis-ci.org/Non-Contradiction/autodiffr)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/Non-Contradiction/autodiffr?branch=master&svg=true)](https://ci.appveyor.com/project/Non-Contradiction/autodiffr)
<!--
[![CRAN_Status_Badge](https://www.r-pkg.org/badges/version/JuliaCall)](https://cran.r-project.org/package=JuliaCall)
[![](https://cranlogs.r-pkg.org/badges/JuliaCall)](https://cran.r-project.org/package=JuliaCall)
[![](https://cranlogs.r-pkg.org/badges/grand-total/JuliaCall)](https://cran.r-project.org/package=JuliaCall)
-->

Package `autodiffr` provides an `R` wrapper for `Julia` packages `ForwardDiff.jl` and `ReverseDiff.jl` to do automatic differentiation for native `R` functions
and some `Rcpp` functions.

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

## Installation

Pakcage `autodiffr` is not on CRAN yet. You can get the development version of `autodiffr` by
```{r eval=FALSE}
devtools::install_github("Non-Contradiction/autodiffr")
```

It's recommended at this stage to also use the development version of `JuliaCall` by
```{r eval=FALSE}
devtools::install_github("Non-Contradiction/JuliaCall")
```

## Basic Usage
```{r}
library(autodiffr)

## Do initial setup

ad_setup()

## If you want to use a julia at a specific location, you could do the following:
## ad_setup(JULIA_HOME = "the folder that contains julia binary"), 
## or you can set JULIA_HOME in command line environment or use `options(...)`

f <- function(x) sum(x^2L)

grad(f, c(2, 3))

g <- grad(f)

g(c(2, 3))
```

## Trouble Shooting and Way to Get Help

### Julia is not found

Make sure the `Julia` installation is correct.
`autodiffr` is able to find `Julia` on PATH,
and there are three ways for `autodiffr` to find `Julia` not on PATH.

- Use `ad_setup(JULIA_HOME = "the folder that contains julia binary")`
- Use `options(JULIA_HOME = "the folder that contains julia binary")`
- Set `JULIA_HOME` in command line environment.

### How to Get Help

- The GitHub Pages for this repository host the documentation for the development
version of `autodiffr`: <https://non-contradiction.github.io/autodiffr/>.

- And you are more than welcome to contact me about `JuliaCall`
at <lch34677@gmail.com> or <cxl508@psu.edu>.

## Suggestion and Issue Reporting

`autodiffr` is under active development now.
Any suggestion or issue reporting is welcome!
You may report it using the link: <https://github.com/Non-Contradiction/autodiffr/issues/new>.
Or email me at <lch34677@gmail.com> or <cxl508@psu.edu>.

## Coding Plan

For coding plan of `autodiffr`, see [Coding Plan](https://github.com/Non-Contradiction/autodiffr/blob/master/Coding%20Plan.md).
