---
title: "Automatic Differentiation in R by autodiffr"
author: "Changcheng Li, John Nash, Hans Werner Borchers"
date: "2018/8/10"
output: github_document
bibliography: opt.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Package `autodiffr` provides an R wrapper for Julia packages [`ForwardDiff.jl`](https://github.com/JuliaDiff/ForwardDiff.jl) and [`ReverseDiff.jl`](https://github.com/JuliaDiff/ReverseDiff.jl) through R package  [`JuliaCall`](https://github.com/Non-Contradiction/JuliaCall) to do **automatic differentiation** for native R functions.
This vignette will walk you through several examples in using `autodiffr`.

## Basic usage

When loading the library `autodiffr` it is still necessary to create a
process running Julia, to connect to it, to check whether all needed
Julia packages are installed resp. to install them in case they are
missing. All this is done by the `ad_setup` function. If everything is
in place, it will finish in short time, otherwise it may take several
minutes (the first time it is called).

    library(autodiffr)

    ad_setup()
    ## Julia version 0.6.4 at location [...]/julia/bin will be used.
    ## Loading setup script for JuliaCall...
    Finish loading setup script for JuliaCall.
    INFO: Precompiling module DiffResults.
    INFO: Precompiling module ForwardDiff.
    INFO: Precompiling module ReverseDiff.

```{r ad_setup, include=FALSE}
library(autodiffr)

ad_setup()

## If julia is not in the path, then autodiffr may fail to locate julia
## in this case, use `ad_setup(JULIA_HOME = "the file folder contains the julia binary")
```

The first time `ad_setup()` is called on your system, these three
Julia modules / packages are being precompiled, later on this will not
happen again, except when the julia packages get installed anew (e.g.,
a new version has appeared). Please note that `ad_setup` is required
whenever you load the `autodiffr` package into R.
And if Julia is not in the path, then `autodiffr` may fail to locate Julia.
In this case, use `ad_setup(JULIA_HOME = "the file folder contains the julia binary")`.

Now automatic differentiation is ready to get applied. 
`autodiffr` has `ad_grad`, `ad_jacobian` and `ad_hessian` to calculate
gradient, jacobian and hessian, 
and has `makeGradFunc`, `makeJacobianFunc` and `makeHessianFunc` to create
gradient, jacobian and hessian functions respectively.
We can see the basic usage below:

```{r basic}
## Define a target function with vector input and scalar output
f <- function(x) sum(x^2L)

## Calculate gradient of f at [2,3] by
ad_grad(f, c(2, 3)) ## deriv(f, c(2, 3))

## Get a gradient function g
g <- makeGradFunc(f)

## Evaluate the gradient function g at [2,3]
g(c(2, 3))

## Calculate hessian of f at [2,3] by
ad_hessian(f, c(2, 3))

## Get a hessian function h
h <- makeHessianFunc(f)

## Evaluate the hessian function h at [2,3]
h(c(2, 3))

## Define a target function with vector input and vector output
f <- function(x) x^2

## Calculate jacobian of f at [2,3] by
ad_jacobian(f, c(2, 3))

## Get a jacobian function j
j <- makeJacobianFunc(f)

## Evaluate the gradient function j at [2,3]
j(c(2, 3))
```

## What to do if autodiffr cannot handle your function?

`autodiffr` works by ultilizing R's S3 object system:
most of R functions are just compositions of many small operations like `+` and `-`,
and these small operations can be dispatched by the S3 object system to Julia,
which finishes the real work of automatic differentiation behind the scence.
This mechanism can work for many R functions,
but it also means that `autodiffr` has same limitations just as R's S3 object system.
For example, `autodiffr` can't handle R's internal C code or some external C and Fortran code.
In this section, we will talk about how to detect to deal with such problems.
We will first talk about the common issues in the next subsection,
and then we will see a little example,
finally we will show how to use a helper function `ad_variant` provided by `autodiffr` to deal with these common problems at the same time.

### Basic R functions that have problems with autodiffr
                       
This is a list of the most common basic R functions that have problems with `autodiffr`:

* `colSums`, `colMeans`, `rowSums`, `rowMeans`: These common functions are all implemented in R's internal C code. And `autodiffr` provides helper functions `cSums`, `cMeans`, `rSums` and `rMeans` to substitute them. Example error message is like "Error in colSums(...) : 'x' must be numeric". 
* `%*%`: this function can't be handled with R's S3 system, so `autodiffr` provides `%m%` to use instead. Example error message is like "Error in x %*% y : requires numeric/complex matrix/vector arguments".
* `crossprod`, `tcrossprod`: these functions are also implemented in R's internal C code, users may use `t(x) %m% y` or `x %m% t(y)` instead, or find some other more effective alternatives. Example error message is like "Error in crossprod(x, y) : requires numeric/complex matrix/vector arguments".
* `diag`: it will also invoke R's internal C code when using `diag` on a vector to create a diagonal matrix and `autodiffr` provides helper functions `diagm` for the same purpose. Example error message is like "Error in diag(x) : long vectors not supported yet: ../../../../R-3.5.1/src/main/array.c:2178".
* `mapply`, `sapply`, `apply`: `autodiffr` provides a helpful function *map*, which can be used similarly to *mapply*.
* `matrix`: it's problematic to use `matrix` to reshape vectors into matrices with `autodiffr`, use R's `array` instead or use the helper function `julia_array` from `autodiffr`. Example error message is like "Error in matrix(x,  :  'data' must be of a vector type, was 'environment'".
*`[<-`: if you ever want to do something like `x[i] <- a` or `x[i, j] <- a`, when `x` is a vector or matrix involved in the differentiation process, then it's likely that you hit this problem: "Error in x[i] <- y : incompatible types (from environment to double) in subassignment type fix". To deal with this problem, maybe you need to do `x <- julia_array(x)` first.

### A little example to demonstrate issues

Let us first define a makeup function to show some of the aforementioned issues:
```{r fun}
fun <- function(x) {
    if (length(x) != 4L) stop("'x' must be a vector of 4 elements.")
    y <- rep(0, 4)
    y[1] <- x[1]; y[2] <- x[2]^2
    y[3] <- x[3]; y[4] <- x[4]^3
    y <- matrix(y, 2, 2)
    det(y)
}
```

If we directly use `autodiffr` to deal with `fun`, we will have:

```{r issue}
x0 <- c(1.2, 1.4, 1.6, 1.8)

try(ad_grad(fun, x0))

## from the error message
## Error in y[1] <- x[1] : 
##  incompatible types (from environment to double) in subassignment type fix
## and the error list in the previous subsection, we can know the problem is from assignment, 
## and we can come up with the solution which deal with the issue:

fun1 <- function(x) {
    if (length(x) != 4L) stop("'x' must be a vector of 4 elements.")
    y <- julia_array(rep(0, 4)) 
    ## we need to use julia_array to wrap the vector
    ## or we can do y <- julia_array(0, 4) which is more elegant.
    y[1] <- x[1]; y[2] <- x[2]^2
    y[3] <- x[3]; y[4] <- x[4]^3
    y <- matrix(y, 2, 2)
    det(y)
}

print(try(ad_grad(fun1, x0)))

## and then we have another error, 
## from the error message 
## Error in matrix(y, 2, 2) : 
##   'data' must be of a vector type, was 'environment'
## and the error list in the previous subsection, we can know the problem is from 
## using matrix to reshape the vector y, 
## and we can come up with the solution which deal with the issue:

fun2 <- function(x) {
    if (length(x) != 4L) stop("'x' must be a vector of 4 elements.")
    ## we need to use julia_array to wrap the vector
    ## or we can do y <- julia_array(0, 4) which is more elegant.
    y <- julia_array(rep(0, 4)) 
    y[1] <- x[1]; y[2] <- x[2]^2
    y[3] <- x[3]; y[4] <- x[4]^3
    ## we need to use array to reshape the vector,
    ## or we can do y <- julia_array(y, 2, 2).
    y <- array(y, c(2, 2))
    det(y)
}

## This time we will have the correct solution
ad_grad(fun2, x0)

## and we can use numeric approximation to check our result:
numDeriv::grad(fun2, x0)
```

### Use `ad_variant` to deal with these issues "automatically"

The above process is a little tedious.
Helpfully, `autodiffr` provides a helper function `ad_variant` to deal with these problems
simutaneously.
And the usage is quite simple and self-explanotory:

```{r ad_variant}
funvariant <- ad_variant(fun)

ad_grad(funvariant, x0)

funvariant1 <- ad_variant(fun, checkArgs = x0)

ad_grad(funvariant1, x0)
```

## A more complex example: Chebyquad function

This problem was given prominence in the optimization literature by @Fletcher65.

First let us have our Chebyquad function. Note that this is for the **vector** x.
This function is taken from the package [`funconstrain`](https://github.com/jlmelville/funconstrain).

```{r cheby}
cyq.f <- function (par) 
{
    n <- length(par)
    if (n < 1) {
        stop("Chebyquad: n must be positive")
    }
    y <- 2 * par - 1
    t0 <- rep(1, n)
    t1 <- y
    ti <- t1
    fsum <- 0
    for (i in 1:n) {
        if (i > 1) {
            ti <- 2 * y * t1 - t0
            t0 <- t1
            t1 <- ti
        }
        fi <- sum(ti)/n
        if (i%%2 == 0) {
            fi <- fi + 1/(i * i - 1)
        }
        fsum <- fsum + fi * fi
    }
    fsum
}
```

Let us choose a single value for the number of parameters, and for illustration
use $n = 10$. 

```{r}
## cyq.setup
n <- 10
x<-1:n
x<-x/(n+1.0) # Initial value suggested by Fletcher
```

For safety, let us check the function and a numerical approximation to the gradient.

```{r}
require(numDeriv)
cat("Initial parameters:")
x
cat("Initial value of the function is ",cyq.f(x),"\n")
gn <- numDeriv::grad(cyq.f, x) # using numDeriv
cat("Approximation to gradient at initial point:")
gn
```

And then we can use `autodiffr` to get the gradient:
```{r}
cyq.ag <- makeGradFunc(cyq.f) # Need autodiffr:: specified for knitr
cat("Gradient at x")
cyq.ag(x)
```

And we can see that the result is very similar, which means that `autodiffr` gives the correct result.

### Create optimized gradient/jacobian/hessian functions with autodiffr

`autodiffr` also provides some ways to generate a more optimized version of gradient/jacobian/hessian functions. See `?autodifffunc`.

Here we will give a simple example for one possible way (and maybe the most effective one) to generate optimized gradient function by `use_tape = TRUE` using the Chebyquad function:
```{r}
cyq.ag_with_tape <- makeGradFunc(cyq.f, x = runif(length(x)), use_tape = TRUE)

cyq.ag_with_tape(x)
```

From the documentation: `use_tape` is whether or not to use tape for reverse mode automatic differentiation. Note that although use_tape can greatly improve the performance sometime, it can only calculate the derivatives w.r.t a certain branching, so it may not give the correct result for functions containing things like if. And this parameter is only effective when mode = "reverse". 

**Important** And also pay attention that for the optimization to have any effect,
`x` must be given and must have the same shape of the potential input(s).
**Important** And also pay attention to the correctness of the optimized gradient function.
Here we can see that the optimized gradient function also gives the correct result.

And then we can use package `microbenchmark` to check the effect of `use_tape` and also compare it with the numeric approximation from `numDeriv`.

```{r}
require(microbenchmark)

cat("cyq.f timing:\n")
tcyq.f <- microbenchmark(cyq.f(x))
tcyq.f

tcyq.ag <- microbenchmark(cyq.ag(x), unit="us" )

tcyq.ag

tcyq.ag_with_tape <- microbenchmark(cyq.ag_with_tape(x), unit="us" )

tcyq.ag_with_tape

tcyq.ng <- microbenchmark(numDeriv::grad(cyq.f, x), unit="us" )

tcyq.ng
```

From the time comparison, we can see that the `use_tape` is very effective and the optimized gradient function performs much better than the numeric approximation provided by `numDeriv` in this case.

## Bibliography
