---
title: "Chebyquad"
author: "Changcheng Li"
date: "2018/7/10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Test problem -- Chebyquad

This problem was given prominence in the optimization literature by @Fletcher65.

First let us define our Chebyquad function. Note that this is for the **vector** x.
This version is expressed as the sum of squares of a vector of function values,
which provide a nonlinear least squares problem. Note that
`crossprod()` may cause difficulties as it is not written in **R**. 


```{r}
require(autodiffr)
ad_setup()
cyq.f <- function (x) {
  rv<-cyq.res(x)
  f <- sum(rv*rv)
}

cyq.res <- function (x) {
# Fletcher's chebyquad function m = n -- residuals 
   n<-length(x)
   res<-zeros(x) # need to use zeros() to do initialization instead of rep() otherwise autodiffr will break.
   ## This is because later on res[i] <- rr, 
   ## if res is a normal R vector and rr is some Julia Number used by autodiffr, 
   ## then R doesn't know what to do.
   for (i in 1:n) { #loop over resids
     rr<-0.0
     for (k in 1:n) {
	z7<-1.0
	z2<-2.0*x[k]-1.0
        z8<-z2
        j<-1
        while (j<i) {
            z6<-z7
            z7<-z8
            z8<-2*z2*z7-z6 # recurrence to compute Chebyshev polynomial
            j<-j+1
        } # end recurrence loop
        rr<-rr+z8
      } # end loop on k
      rr<-rr/n
      if (2*trunc(i/2) == i) { rr <- rr + 1.0/(i*i - 1) }
      res[i]<-rr
    } # end loop on i
    res
}
```

Let us choose a single value for the number of parameters, and for illustration
use $n = 4$. 

```{r}
## cyq.setup
n <- 4
  lower<-rep(-10.0, n)
  upper<-rep(10.0, n) 
  x<-1:n
  x<-x/(n+1.0) # Initial value suggested by Fletcher
```

For safety, let us check the function and a numerical approximation to the gradient.

```{r}
require(numDeriv)
cat("Initial parameters:")
print(x)
cat("Initial value of the function is ",cyq.f(x),"\n")
gn <- numDeriv::grad(cyq.f, x) # using numDeriv
cat("Approximation to gradient at initial point:")
print(gn)
```

Using a modular approach to the problem, first specifying it
via **residuals** and computing the function as a sum of squares, we can
also generate the gradient. 

```{r}
# Ref: Fletcher, R. (1965) Function minimization without calculating derivatives -- a review,
#         Computer J., 8, 33-41.

# Note we do not have all components here e.g., .jsd, .h

cyq.jac<- function (x) {
#  Chebyquad Jacobian matrix
   n<-length(x)
   cj<-matrix(0.0, n, n)
   for (i in 1:n) { # loop over rows
     for (k in 1:n) { # loop over columns (parameters)
       z5<-0.0
       cj[i,k]<-2.0
       z8<-2.0*x[k]-1.0 
       z2<-z8
       z7<-1.0
       j<- 1
       while (j<i) { # recurrence loop
         z4<-z5
         z5<-cj[i,k]
         cj[i,k]<-4.0*z8+2.0*z2*z5-z4
         z6<-z7
         z7<-z8
         z8<-2.0*z2*z7-z6
         j<- j+1
       } # end recurrence loop
       cj[i,k]<-cj[i,k]/n
     } # end loop on k
   } # end loop on i
   cj
}


cyq.g <- function (x) {
   cj<-cyq.jac(x)
   rv<-cyq.res(x)
   gg<- 2.0 * as.vector(rv %*% cj)
}


# check gradient function cyq.g

gajn <- cyq.g(x)
print(gajn)
```
We can now try to see if `autodiffr` matches this gradient. 
However, the following code gives an error in Julia.

```{r, eval=TRUE}
# Do not evaluate, as this fails # Now it should work
 cyq.ag <- autodiffr::makeGradFunc(cyq.f)
gaag <- cyq.ag(x)
print(gaag)
```

As a workaround, we can get the Chebyquad function from the package `funconstrain`.
The `funconstrain`
offering does NOT require a call to the residuals, but has a single
level **R** function.

```{r, eval=TRUE, echo=TRUE}
require(funconstrain)
cat("funconstrain loaded\n")
cheb <- chebyquad() # Seem to need the brackets or doesn't return pieces
print(str(cheb))
cyq2.f <- cheb$fn
## Note that funconstrain offers the starting value
## x0b <- cheb$x0(n=4) # Need the size of the vector
## x0b
## cyq2.f(x0b)
## same as
print(cyq2.f(x))
## Try the gradient
cyq2.ag <- autodiffr::makeGradFunc(cyq2.f) # Need autodiffr:: specified for knitr
## print(cyq2.g)
cat("Gradient at x")
g2ag <- cyq2.ag(x)
print(g2ag)

require(microbenchmark)

cat("cyq.f timing:\n")
tcyq.f <- microbenchmark(cyq.f(x))
tcyq.f
cat("cyq2.f timing:\n")
tcyq2.f <- microbenchmark(cyq2.f(x))
tcyq2.f
tcyq.g <- microbenchmark(cyq.g(x))
tcyq.g
tcyq.g <- microbenchmark(cyq.g(x))
tcyq.g

cyq2.g <- cheb$gr
tcyq2.g <- microbenchmark(cyq2.g(x), unit="us" )
# microseconds
tcyq2.g

# These are very slow
tcyq.ag <- microbenchmark(cyq.ag(x), unit="us" )
# microseconds
tcyq.ag
tcyq2.ag <- microbenchmark(cyq2.ag(x), unit="us" )
# microseconds
tcyq2.ag

# These are quicker, but still slow
cyq.optimized_ag <- autodiffr::makeGradFunc(cyq.f, x = runif(length(x)), use_tape = TRUE)
cyq2.optimized_ag <- autodiffr::makeGradFunc(cyq2.f, x = runif(length(x)), use_tape = TRUE)
tcyq.optimized_ag <- microbenchmark(cyq.optimized_ag(x), unit="us" )
# microseconds
tcyq.optimized_ag
tcyq2.optimized_ag <- microbenchmark(cyq2.optimized_ag(x), unit="us" )
# microseconds
tcyq2.optimized_ag

## The slowness of the optimized method is partly because the overhead of the JuliaCall and autodiffr package.
## After interface functions become stable, I will try to carry on some performance optimizations,
## which is a goal of the project at last phase.
## For example, even we only have a very simple function, the timing is high because of the overhead.
foobar <- function(x) sum(x)
foobar.ag <- autodiffr::makeGradFunc(foobar, x = runif(length(x)), use_tape = TRUE)
tfoobar.ag <- microbenchmark(foobar.ag(x), unit="us" )
tfoobar.ag

## Suppose we are dealing with input of larger size, the overhead stays roughly the same,
## so the overhead should matters not that much as in the case n = 4.
## For example, if n = 25, the diffrence of performance in ratio is not that much.
## cyq.setup
n <- 25
lower<-rep(-10.0, n)
upper<-rep(10.0, n) 
x<-1:n
x<-x/(n+1.0) # Initial value suggested by Fletcher

tcyq.g <- microbenchmark(cyq.g(x), unit = "us")
tcyq.g

tcyq2.g <- microbenchmark(cyq2.g(x), unit="us" )
# microseconds
tcyq2.g

## The bad thing is that if the input size changes, we need to make an optimized gradient again.
cyq.optimized_ag <- autodiffr::makeGradFunc(cyq.f, x = runif(length(x)), use_tape = TRUE)
cyq2.optimized_ag <- autodiffr::makeGradFunc(cyq2.f, x = runif(length(x)), use_tape = TRUE)
tcyq.optimized_ag <- microbenchmark(cyq.optimized_ag(x), unit="us" )
# microseconds
tcyq.optimized_ag
tcyq2.optimized_ag <- microbenchmark(cyq2.optimized_ag(x), unit="us" )
# microseconds
tcyq2.optimized_ag

## Some benchmarking based on the new debugged argument.
cyq.optimized_ag_debugoff <- autodiffr::makeGradFunc(cyq.f, x = runif(length(x)), use_tape = TRUE, debugged = FALSE)
cyq2.optimized_ag_debugoff <- autodiffr::makeGradFunc(cyq2.f, x = runif(length(x)), use_tape = TRUE, debugged = FALSE)
tcyq.optimized_ag_debugoff <- microbenchmark(cyq.optimized_ag_debugoff(x), unit="us" )
# microseconds
tcyq.optimized_ag_debugoff
tcyq2.optimized_ag_debugoff <- microbenchmark(cyq2.optimized_ag_debugoff(x), unit="us" )
# microseconds
tcyq2.optimized_ag_debugoff

## Also note it is better to check the correctness when generating optimized gradient,
all.equal(cyq.g(x), cyq.optimized_ag(x))
all.equal(cyq.g(x), cyq2.optimized_ag(x))

all.equal(cyq.g(x), cyq.optimized_ag_debugoff(x))
all.equal(cyq.g(x), cyq2.optimized_ag_debugoff(x))

## Benchmarking times without user interface wrappers
tape1 <- reverse.grad.tape(cyq.f, runif(length(x)))
microbenchmark(reverse.grad(tape1, x), unit="us" )

tape2 <- reverse.grad.tape(cyq2.f, runif(length(x)))
microbenchmark(reverse.grad(tape2, x), unit="us" )

## Benchmarking times without autodiffr wrappers
JuliaCall::julia_command("using ReverseDiff")
tape1 <- reverse.grad.tape(cyq.f, runif(length(x)))
microbenchmark(JuliaCall::julia_call("ReverseDiff.gradient!", tape1, x), unit="us" )

tape2 <- reverse.grad.tape(cyq2.f, runif(length(x)))
microbenchmark(JuliaCall::julia_call("ReverseDiff.gradient!", tape2, x), unit="us" )
```
